---
title: Acute Virus Infection and Immune Response Model
output:
  html_document:
    theme: null
    highlight: null
    fig_retina: null
    fig_caption: true
    mathjax: default 
    keep_md: false
bibliography: dsairm_references.bib
---

```{r, include = FALSE}
#load various variable definitions that are the same for each app
source('startup_script.R')
currentrmdfile = knitr::current_input() 
appsettings = get_settings(currentrmdfile,appdocdir,packagename)
```



## Overview {#shinytab1}
This app allows exploration of an acute virus infection model, with compartments for uninfected cells, infected cells and (free) virus. The model also includes abstract versions of the innate and adaptive immune response. Read about the model in *The Model* tab. Then, work through the tasks described in the *What To Do* tab.

The model used here is a variation of the one introduced in the ___Basic Virus Model___ app. If you haven't done so, check out and explore that app first. 


### Learning Objectives

* Learn about acute virus infection dynamics.
* Understand the impact the immune response can have on outcomes in this model.


## The Model {#shinytab2}

### Model Overview
This model consists of 5 compartments and can capture some of the basic dynamics of viral infections and the ensuing immune response. In this model, we track the following entities, by assigning each to a compartment:

* **U** - uninfected cells 
* **I** - infected cells
* **V** - (free) virus
* **F** - innate (interferon) response
* **T** - adaptive (T-cell) response


In addition to specifying the ***compartments*** of a model, we need to specify the dynamics determining the changes for each compartment. Broadly speaking, there are ***processes*** that increase the numbers in a given compartment/stage, and processes that lead to a reduction. Those processes are sometimes called in-flows and out-flows. 

For our system, we specify the following processes/flows: 

* Virus infects cells at rate _b_.
* Infected cells produce new virus at rate _p_ and die at rate _d~I~_.
* Free virus is removed at rate _d~V~_ or goes on to infect further uninfected cells at rate _gb_. To allow conversion from infectious virus units in the model to some experimental units (e.g. plaque forming units), an additional conversion factor, _g_, is included in the model.
* The innate (interferon) response is induced proportional to the number of infected cells at rate _r~F~_ and decays at rate _d~F~_. The innate response lowers the rate of virus production by infected cells, with strength of reduction controlled by parameter _k_.
* The adaptive (CD8 T-cell) response grows exponentially at rate _r~T~_ and proportional to the strength of the innate response, and decays at rate _d~T~_. T-cells kill infected cells at rate _k~T~_.


### Model Diagram
The diagram illustrating this compartmental model is shown in the figure.

```{r modeldiagram,  fig.cap='Model Diagram',  echo=FALSE, out.width = "70%"}
knitr::include_graphics( system.file(figuredir,appsettings$modelfigname,package=packagename))
```


### Model Equations
Implementing this model as a ***continuous-time, deterministic model*** leads to the following set of ***ordinary differential equations***. 


\begin{aligned}
\dot U & = - b U V  \\
\dot I & = b U V - d_I I - k_T T I   \\
\dot V & = \frac{p}{1+k_F F}I - d_V V - gb UV \\
\dot F & = r_F I - d_F F \\
\dot T & = r_T T F - d_T T
\end{aligned}





## What To Do {#shinytab3}

**The tasks below are described in a way that assumes everything is in units of days (rate parameters, therefore, have units of inverse days). If any quantity is not given in those units, you need to convert it first (e.g. if it says a week, you need to convert it to 7 days).**



```{r, echo=FALSE, eval=TRUE}
#this is the running counter for the records which starts at 1 
rc=1

#empty object, will hold all outcomes
alloutcomes = NULL

#########################
# Task 1
#########################
tid = 1
tasktext = "Let's start by considering infection dynamics in the absence of the immune response. Set the initial conditions to 10^5^ uninfected cells, no infected cells, and 10 virus particles. Assume also that infected cells have an average life-span of 1 day, and virus has a life-span of 6 hours (remember that the inverse of the lifespan is the rate of death, and make sure you convert to the right units). Set that the virus production by an infected cell is 100 virions per day and that the rate at which new cells become infected is 10^-6^. Assume there is no need to do any unit adjustment/conversion (i.e. the value of that parameter is 1). Set the starting values for the immune response variables ~F~ and ~T~ to 0, and also set the two immune response induction rates to 0. With those choices, the values for the other immune response related parameters do not matter. If you want, you can also set them all to 0. \nRun the simulation for 50 days; produce plots with and without log-scales. You should get a single, acute infection with virus and infected cells rising at first and then declining. At the end you should be left with around 11069 uninfected cells and essentially no infected cells or virus."
nrec = 1 # number of items to record
out_records = c("Number of infected cells at the peak of infection")
out_types = rep("Rounded_Integer",nrec)
out_notes = rep("Report the rounded integer",nrec)
outcomes = data.frame( TaskID = rep(tid,nrec),
                       TaskText = rep(tasktext,nrec),
                      RecordID = paste0('T',tid,'R',(1:nrec)),
                      Record = out_records, 
                      Type = out_types, 
                      Note = out_notes)
alloutcomes = rbind(alloutcomes,outcomes)
rc = rc + nrec #increment record counter by number of outcomes to record for this task 

#########################
# Task 2
#########################
tid = tid + 1
tasktext = "In the absence of the immune response, this model is the same as the one covered in the first part of the _Basic Virus_ app. Convince yourself that you can get at most one rise, then decline of virus, and confirm that this acute infection increases as you increase virus production and infection rates, and that it decreases as you increase infected and virus death rates. \nRun a simulation for 50 days with a doubling of virus production and infection rate (compared to the previous task), as well as a doubling of the lifespans of the infected cells and virus (remember the relation between durations, such as lifespan, and rates, such as death rates.)"
nrec = 1 # number of items to record
out_records = c("Remaining infected cells at the end")
out_types = rep("Rounded_Integer",nrec)
out_notes = rep("Report the rounded integer",nrec)
outcomes = data.frame( TaskID = rep(tid,nrec),
                       TaskText = rep(tasktext,nrec),
                      RecordID = paste0('T',tid,'R',(1:nrec)),
                      Record = out_records, 
                      Type = out_types, 
                      Note = out_notes)
alloutcomes = rbind(alloutcomes,outcomes)
rc = rc + nrec #increment record counter by number of outcomes to record for this task 


#########################
# Task 3
#########################
tid = tid + 1
tasktext = "Now we'll turn on the immune response. Set everything as you did for the first task. Then change the starting value for ~T~ to 1, set ~r_T_~ = 0.01, ~d_T~ = 1 and ~k_T_~ = 0.01.  Run the simulation and convince yourself that in the absence of the innate response, the adaptive response is not triggered, and the resulting infection is thus like the one you saw in the previous tasks."
nrec = 1 # number of items to record
out_records = c("Remaining uninfected cells at the end")
out_types = rep("Rounded_Integer",nrec)
out_notes = rep("Report the rounded integer",nrec)
outcomes = data.frame( TaskID = rep(tid,nrec),
                       TaskText = rep(tasktext,nrec),
                      RecordID = paste0('T',tid,'R',(1:nrec)),
                      Record = out_records, 
                      Type = out_types, 
                      Note = out_notes)
alloutcomes = rbind(alloutcomes,outcomes)
rc = rc + nrec #increment record counter by number of outcomes to record for this task 

#########################
# Task 4
#########################
tid = tid + 1
tasktext = "The previous task showed that in the absence of the innate response, the adaptive response won't kick in. The reverse is not true. Set everything back as in task 1. Then set the starting value for ~F~ to 1, ~r_F_~ = 0.01, ~d_F~ = 1 and ~k_F_~ = 0. With these settings, the innate response is activated, but has no impact yet on the virus kinetics. You should see the same virus and infected cell trajectories as for task 1. Now, turn on the action of the innate response by setting ~k_F_~ = 0.01."
outcomes = data.frame( TaskID = rep(tid,nrec),
                       TaskText = rep(tasktext,nrec),
                      RecordID = paste0('T',tid,'R',(1:nrec)),
                      Record = out_records, 
                      Type = out_types, 
                      Note = out_notes)
alloutcomes = rbind(alloutcomes,outcomes)
rc = rc + nrec #increment record counter by number of outcomes to record for this task 


```



```{r echo=FALSE}
#save the fully filled task table to a tsv file
alloutcomes$QuizID = paste0(packagename,"_",appsettings$appid)
alloutcomes$AppTitle = appsettings$apptitle
alloutcomes$AppID = appsettings$appid
#remove a few variables from the data frame
savedoutcomes <- dplyr::select(alloutcomes,QuizID,AppID,AppTitle,TaskID,TaskText,RecordID,Record,Type,Note)		
write.table(savedoutcomes, paste0(appsettings$appid,"_tasktable.tsv"), append = FALSE, sep = "\t", row.names = F, col.names = TRUE)
```


```{r, echo=FALSE, results='asis'}
# Take all the text stored in the table and print the tasks and items to record
write_tasktext(alloutcomes)
```




## Further Information {#shinytab4}

This app (and all others) are structured such that the Shiny part (the graphical interface you see and the server-side function that goes with it) calls an underlying R script (or several) which runs the simulation for the model of interest and returns the results.
  
For this app, the underlying function running the simulation is called ``r appsettings$simfunction``. You can call them directly, without going through the shiny app. Use the `help()` command for more information on how to use the functions directly. If you go that route, you need to use the results returned from this function and produce useful output (such as a plot) yourself. 
  
You can also download all simulator functions and modify them for your own purposes. Of course to modify these functions, you'll need to do some coding. For examples on using the simulators directly and how to modify them, read the package vignette by typing `vignette('DSAIRM')` into the R console.
  
This model is based on one of the models discussed in [@li14]. See also for instance [@beauchemin11; @smith11] and references in the other apps in this section. 


### References


