---
title: Chronic Virus Infection and Immune Response Model
output:
  html_document:
    theme: null
    highlight: null
    fig_retina: null
    fig_caption: true
    mathjax: default 
    keep_md: false
bibliography: dsairm_references.bib
---

```{r, include = FALSE}
#load various variable definitions that are the same for each app
source('startup_script.R')
currentrmdfile = knitr::current_input() 
appsettings = get_settings(currentrmdfile,appdocdir,packagename)
```


## Overview {#shinytab1}
This app allows exploration of a chronic virus infection model, with compartments for uninfected cells, infected cells and (free) virus. The model also includes abstract versions of the innate and adaptive immune response. Read about the model in *The Model* tab. Then, work through the tasks described in the *What To Do* tab.


### Learning Objectives

* Learn about chronic virus infection dynamics.
* Understand the impact the immune response can have on outcomes in this model.


## The Model {#shinytab2}

### Model Overview
This model consists of 5 compartments and can capture some of the basic dynamics of viral infections and the ensuing immune response. In this model, we track the following entities, by assigning each to a compartment:

* **U** - uninfected cells 
* **I** - infected cells
* **V** - (free) virus
* **F** - innate (interferon) response
* **T** - adaptive (T-cell) response


In addition to specifying the ***compartments*** of a model, we need to specify the dynamics determining the changes for each compartment. Broadly speaking, there are ***processes*** that increase the numbers in a given compartment/stage, and processes that lead to a reduction. Those processes are sometimes called in-flows and out-flows. 

For our system, we specify the following processes/flows: 

* Uninfected cells are produced at some rate _n_ and naturally die at some rate _d~U~_.
* Virus infects cells at rate _b_.
* Infected cells produce new virus at rate _p_ and die at rate _d~I~_.
* Free virus is removed at rate _d~V~_ or goes on to infect further uninfected cells at rate _gb_. To allow conversion from infectious virus units in the model to some experimental units (e.g. plaque forming units), an additional conversion factor, _g_, is included in the model.
* The innate (interferon) response is induced proportional to the number of infected cells at rate _r~F~_ and decays at rate _d~F~_. The innate response lowers the rate of virus production by infected cells, with strength of reduction controlled by parameter _k_.
* The adaptive (CD8 T-cell) response grows exponentially, proportional to the virus load at rate _r~T~_ and decays at rate _d~T~_. T-cells kill infected cells at rate _k~T~_. 

Note that the choice of growth dynamics is slightly different from the acute virus model. There, we assumed the growth was proportional to the innate response ~F~. Here, we assume it is driven by virus/antigen load, ~V~. Both assumptions are reasonable approximations. In reality, it is likely a complex interaction between virus/antigen load and innate response that drives adaptive responses. Trying to model this in detail is beyond the goal of this app. We want to keep it reasonably simple here so we can get some intuition in these kinds of models. For any real problem, you need to decide which components and processes to include and how to model them.


### Model Diagram
The diagram illustrating this compartmental model is shown in the figure.

```{r modeldiagram,  fig.cap='Model Diagram',  echo=FALSE, out.width = "70%"}
knitr::include_graphics( system.file(figuredir,appsettings$modelfigname,package=packagename))
```


### Model Equations
Implementing this model as a ***continuous-time, deterministic model*** leads to the following set of ***ordinary differential equations***. 


\begin{aligned}
\dot U & = n - d_U U - b U V  \\
\dot I & = bUV - d_I I - k_T T I   \\
\dot V & = \frac{p}{1+k_F F}I - d_V V - gb UV \\
\dot F & = r_F I - d_F F \\
\dot T & = r_T T V - d_T T
\end{aligned}




## What To Do {#shinytab3}

**The tasks below are described in a way that assumes everything is in units of days (rate parameters, therefore, have units of inverse days). If any quantity is not given in those units, you need to convert it first (e.g. if it says a week, you need to convert it to 7 days).**



```{r, echo=FALSE, eval=TRUE}
#this is the running counter for the records which starts at 1 
rc=1

#empty object, will hold all outcomes
alloutcomes = NULL

#########################
# Task 1
#########################
tid = 1
tasktext = "Let's start by considering an acute viral infection. Set the initial conditions to 10^5^ uninfected cells, no infected cells, and 10 virus particles. We make the assumption that on the timescale of an acute infection (several days), the processes of natural, uninfected cell turnover are so slow that they can be ignored. Set values for the uninfected cell birth and death rates to reflect this assumption. Assume also that infected cells have an average life-span of 1 day, and virus has a life-span of 6 hours (remember that the inverse of the lifespan is the rate of death, and make sure you convert to the right units). Set that the virus production by an infected cell is 100 virions per day and that the rate at which new cells become infected is 10^-6^. Assume there is no need to do any unit adjustment/conversion (i.e. the value of that parameter is 1). 

\nRun the simulation for 50 days; produce plots with and without log-scales. You should get a single, acute infection with virus and infected cells rising at first and then declining. At the end you should be left with around 11069 uninfected cells and no infected cells or virus."
nrec = 1 # number of items to record
out_records = c("Number of infected cells at the peak of infection")
out_types = rep("Rounded_Integer",nrec)
out_notes = rep("Report the rounded integer",nrec)
outcomes = data.frame( TaskID = rep(tid,nrec),
                       TaskText = rep(tasktext,nrec),
                      RecordID = paste0('T',tid,'R',(1:nrec)),
                      Record = out_records, 
                      Type = out_types, 
                      Note = out_notes)
alloutcomes = rbind(alloutcomes,outcomes)
rc = rc + nrec #increment record counter by number of outcomes to record for this task 

#########################
# Task 2
#########################
tid = tid + 1
tasktext = "Slowly increase the virus death rate in increments of 0.5. Contemplate what you expect to see, then run the simulation to compare. Keep increasing until you get essentially no more infection (i.e., < 1% initial uninfected cells become infected). You will have to adjust the simulation time for that, too."
nrec = 2 # number of items to record
out_records = c("Virus death rate at which no infection occurs", "Virus lifespan (in days) corresponding to the death rate at which no infection occurs")
out_types = rep("Numeric", nrec)
out_notes = rep("Round to two significant digits, report as non-scientic notation (as X.YZ)",nrec)
outcomes = data.frame( TaskID = rep(tid,nrec),
                       TaskText = rep(tasktext,nrec),
                      RecordID = paste0('T',tid,'R',(1:nrec)),
                      Record = out_records, 
                      Type = out_types, 
                      Note = out_notes)
alloutcomes = rbind(alloutcomes,outcomes)
rc = rc + nrec #increment record counter by number of outcomes to record for this task 

```



```{r echo=FALSE}
#save the fully filled task table to a tsv file
alloutcomes$QuizID = paste0(packagename,"_",appsettings$appid)
alloutcomes$AppTitle = appsettings$apptitle
alloutcomes$AppID = appsettings$appid
#remove a few variables from the data frame
savedoutcomes <- dplyr::select(alloutcomes,QuizID,AppID,AppTitle,TaskID,TaskText,RecordID,Record,Type,Note)		
write.table(savedoutcomes, paste0(appsettings$appid,"_tasktable.tsv"), append = FALSE, sep = "\t", row.names = F, col.names = TRUE)
```


```{r, echo=FALSE, results='asis'}
# Take all the text stored in the table and print the tasks and items to record
write_tasktext(alloutcomes)
```




## Further Information {#shinytab4}

This app (and all others) are structured such that the Shiny part (the graphical interface you see and the server-side function that goes with it) calls an underlying R script (or several) which runs the simulation for the model of interest and returns the results.
  
For this app, the underlying function running the simulation is called ``r appsettings$simfunction``. You can call them directly, without going through the shiny app. Use the `help()` command for more information on how to use the functions directly. If you go that route, you need to use the results returned from this function and produce useful output (such as a plot) yourself. 
  
You can also download all simulator functions and modify them for your own purposes. Of course to modify these functions, you'll need to do some coding. For examples on using the simulators directly and how to modify them, read the package vignette by typing `vignette('DSAIRM')` into the R console.
  
If you want to learn a bit more about these kinds of models applied to acute viral infections, specifically influenza, see e.g. [@beauchemin11; @smith11]. A few examples of these kinds of models applied to chronic viral infections, see e.g. [@guedj10; @chatterjee12; @perelson13]. 


### References


