---
title: Uncertainty and Sensitivity Analysis
output:
  html_document:
    theme: null
    highlight: null
    fig_retina: null
    fig_caption: true
    mathjax: default 
    keep_md: false
bibliography: dsairm_references.bib
---


```{r, include = FALSE}
#load various variable definitions that are the same for each app
source('startup_script.R')
sapply(files_to_source, source) #source some helper files defined in the files_to_source variable
currentrmdfile = knitr::current_input()  #get current file name
appsettings = get_settings(currentrmdfile,appdocdir,packagename) #get settings for current app
```



## Overview {#shinytab1}
This app allows exploration of the concept of uncertainty and sensitivity analysis. 
For this purpose, we use the _Basic Bacteria Model_ introduced in the app of that name. If you haven't done so yet, familiarize yourself and work through that app first. It might also be a good idea to go through the __Bacteria Model Exploration__ app to familiarize yourself further with the model and the idea of recording outcomes of interest as a function of changing parameters.

### Learning Objectives

* Be familiar with the motivation behind doing an uncertainty and sensitivity (U/S) analysis 
* Know the steps involved in doing an U/S analysis.

## The Model {#shinytab2}

### Model Overview

The underlying model is the continuous-time, ODE version of the model from the  __Basic Bacteria Model__ app. Details can be found there. 


For convenience, here is a quick summary, diagram and equations again.

We model 2 compartments:

* **B** - bacteria 
* **I** - immune response

We specify the following processes/flows: 

* Bacteria grow/divide at some maximum rate (which we label _g_) and saturate as they approach some maximum carrying capacity, _B~max~_. 
* Bacteria die at a natural death rate (which we label _d~B~_).
* Bacteria are killed by the immune response at some rate _k_.
* The immune response grows proportional to the number of bacteria and itself at some rate _r_.
* The immune response decays at some rate (which we label _d~I~_).

The equations are given by:

$$
\begin{aligned}
\dot B & = gB(1-\frac{B}{B_{max}})-d_B B - k BI \\
\dot I & = r B I - d_I I
\end{aligned}
$$

And the flow diagram is as follows

```{r modeldiagram,  fig.cap='Model Diagram',  echo=FALSE, out.width = "70%"}
knitr::include_graphics( system.file(figuredir,appsettings$modelfigname,package=packagename))
```




### Uncertainty and Sensitivity analysis

For the system we want to study and model, we often have only rough estimates for the model parameters and starting values. Instead of specifying fixed values (which results in a single time-series), we can instead specify parameter ranges, choose sets of parameter values from these ranges, and run the model for multiple sets of parameters. 

The simplest way of specifying parameter ranges is to set an upper and lower bound (based on what we know about the biology of the system) and randomly choose any value within those bounds. We can almost always set bounds even if we know very little about a system. Assume we want to model the death rate of some cell type (e.g. NK cells) in humans. We might not know anything, but we can still be fairly confident that their lifespan is at least 1 second and less than 100 years. That's of course a wide range and we should and usually can narrow ranges further, based on biological knowledge of a given system.

If we are fairly certain that values are close to some quantity, instead of specifying a uniform distribution, we can choose one that is more peaked around the most likely value. Normal distributions are not ideal since they allow negative values, which doesn't make sense for our parameters. The gamma distribution is a better idea, since it leads to only positive values.

To run the model for this app, we need to specify values for the 2 initial conditions, _B~0~_ and _I~0~_, and the 6 model parameters _g_, _B~max~_, _d~B~_, _k_, _r_, _d~I~_. All initial conditions and parameters are sampled uniformly between the specified upper and lower bound, apart from the bacteria growth rate, which is given by a gamma distribution, with user-specified mean and variance. For this teaching app, there is no biological reason for making bacterial growth different, I just picked one parameter and decided to make it non-uniformly distributed to show you different ways one can implement distributions from which to draw parameter samples.

The way the samples are drawn could be done completely randomly, but that would lead to inefficient sampling. A smarter method exists, known as Latin Hypercube sampling (LHS). It essentially ensures that we sample the full range of possible parameter combinations in an efficient manner. For more technical details, see e.g. [@saltelli04]. For this app, we use LHS.

Once we specify the ranges for each parameter, the sampling method, and the number of samples, the simulation draws that many samples, runs the model for each sample, and records outcomes of interest. While the underlying simulation returns a time-series for each sample, we are usually not interested in the full time-series. Instead, we are interested in some summary quantity. For instance in this model, we might be interested in the maximum/peak level of bacteria during the infection, the level of bacteria at the end (the steady state) of the infection, and the level of the immune response at steady state. This app records and reports those 3 quantities as _B~peak~_, _B~steady~_ and _I~steady~_.

Results from such simulations for multiple samples can be analyzed in different ways. The most basic one, called **Uncertainty Analysis** only asks what level of uncertainty we have in our outcomes of interest, given the amount of uncertainty in our model parameter values. This can be graphically represented with a boxplot, and is one of the plot options for this app.

In a next step, we can ask 'how sensitive is the outcome(s) of interest to variation in specific parameters' - that part is the **Sensitivity Analysis**. When you run the simulations, you essentially do both uncertainty and sensitivity analysis at the same time, it's just a question of how you further process the results. We can graphically inspect the relation between outcome and some parameter with scatterplots. If we find that there is a monotone up or down (or neither) trend between parameter and outcome, we can also summarize the finding using a correlation coefficient. For this type of analysis, using the Spearman rank correlation coefficient is useful, which is what the app produces below the figures.


### A note on randomness in computer simulations

This simulation (as well as some of the others) involves sampling. This leads to some level of randomness. In science, we want to be as reproducible as possible. Fortunately, random numbers on a computer are not completely random, but can be reproduced. In practice, this is done by specifying a random number seed, in essence a starting position for the algorithm to produce pseudo-random numbers. As long as the seed is the same, the code should produce the same pseudo-random numbers each time, thus ensuring reproducibility.


## What to do {#shinytab3}

First, familiarize yourself with the setup of the app, it looks different from most others. Parameters are not set to specific values. Instead, most parameters have a lower and upper bound. For each simulation that is run, random values for the parameter are chosen  uniformly between those bounds. The parameter _g_ does not have a uniform but instead a gamma distribution, you can specify its mean and variance to determine the distribution from which values are sampled. 

The default outcome plots are boxplots, which show the distribution of the 3 outcomes of interest for the different parameter samples. You can set the number of samples you want to run. Samples are constructed using the latin hypercube method to efficiently span the space of possible parameter values. In general, more samples are better, but of course take longer to run.


```{r, echo=FALSE, eval=TRUE}

#this is the running counter for the records which starts at 1 
rc=1

#empty object, will hold all outcomes
alloutcomes = NULL

#########################
# Task 1
#########################
tid = 1
tasktext = "The creation of parameter samples involves some element of uncertainty, and we need to make use of random numbers. We still want results to be reproducible. That's where the random number seed comes in. As long as the seed is the same, the code should produce the same pseudo-random numbers each time, thus ensuring reproducibility. Let's explore this. Leave all settings at their default values. Run the simulation. You should find that the median for Bsteady is around 10736 (as printed below the plot). Run the simulation again with the same settings, you should get exactly the same results. Now change the random number seed (rngseed) to 101. Run the simulation again. The median value should be lower. Note that the only thing that is different is that the computer drew different random samples for the parameter values. Keep exploring other rngseed values. You should notice that sometimes results change a bit, sometimes a lot, and that it doesn't matter if you change the seed by just a bit or a lot."
nrec = 1 # number of items to record
out_records = c("Median for Bsteady with rngseed = 101.")
out_types = rep("Rounded_Integer",nrec)
out_notes = rep("Report the rounded integer",nrec)
outcomes = data.frame( TaskID = rep(tid,nrec),
                       TaskText = rep(tasktext,nrec),
                      RecordID = paste0('T',tid,'R',(1:nrec)),
                      Record = out_records, 
                      Type = out_types, 
                      Note = out_notes)
alloutcomes = rbind(alloutcomes,outcomes)
rc = rc + nrec #increment record counter by number of outcomes to record for this task 

#########################
# Task 2
#########################
tid = tid + 1
tasktext = "The more samples you have, the more robust the results are to changes in the underlying sample generation (determined by the random number seed). Try checking this by running 100 instead of 10 samples with the two random number seeds from above. Be patient, this might take a good while. Since each sample means a run of the underlying simulation model, things start to go slow for increased sample size. For rngseed = 100, the median is around 9306, for the second rngseed value it is somewhat higher. You might have noticed that the variability shrinks in the central quantities (mean, median) for the larger sample size, not much in the extremes (min/max). You might have also noticed that some Bpeak is inherently more variable than Bsteady and Isteady. Also note the 'system might not have reached steady state' message. If for too many of the samples steady state has not been reached, the results for _B~steady~_ and _I~steady~_ are not correct. In that case you need to increase the simulation time to allow the system to settle into steady state. For some parameter combinations, that can take very long."
nrec = 1 # number of items to record
out_records = c("Median for Bsteady with rngseed = 101, 100 samples.")
out_types = rep("Rounded_Integer",nrec)
out_notes = rep("Report the rounded integer",nrec)
outcomes = data.frame( TaskID = rep(tid,nrec),
                       TaskText = rep(tasktext,nrec),
                      RecordID = paste0('T',tid,'R',(1:nrec)),
                      Record = out_records, 
                      Type = out_types, 
                      Note = out_notes)
alloutcomes = rbind(alloutcomes,outcomes)
rc = rc + nrec #increment record counter by number of outcomes to record for this task 

#########################
# Task 3
#########################
tid = tid + 1
tasktext = "Recall the underlying model and its behavior. If you can't, revisit the __Basic Bacteria__ app. Use your understanding of the model to predict what happens if you increase both lower and upper bound for the immune response activation rate, _r_. Increase both lower and upper bounds (rmin and rmax) by a factor of 10. Set seed to 100, with 50 samples. Leave everything else at the default values. Run simulations, see how results change. Now go the opposite direction, lower the initial lower/upper bounds by a factor of 10 from their original value. Run simulations, see how results change. Especially for the steady states, you know from previous apps how things should change as you change _r_. Compare your simulation results with your expectations."
nrec = 1 # number of items to record
out_records = c("Median for Bsteady rmin = 0.001 and rmax = 0.002.")
out_types = rep("Rounded_Integer",nrec)
out_notes = rep("Report the rounded integer",nrec)
outcomes = data.frame( TaskID = rep(tid,nrec),
                       TaskText = rep(tasktext,nrec),
                      RecordID = paste0('T',tid,'R',(1:nrec)),
                      Record = out_records, 
                      Type = out_types, 
                      Note = out_notes)
alloutcomes = rbind(alloutcomes,outcomes)
rc = rc + nrec #increment record counter by number of outcomes to record for this task 


#########################
# Task 4
#########################
tid = tid + 1
tasktext = "Continue this exploration as you change ranges for different parameters. Note that while you change ranges for some parameters, every time you get samples for all parameters in those ranges. This is different from the __Basic Bacteria Exploration__ app where only the parameter we scanned over changed and every other parameter remainded fixed. Usually, you have a parameter you are most interested in, and you'll explore that parameter in detail, while allowing for variability in the others. As you play around, it is likely that for some settings you'll see warning or error messages on the `R` console. That generally means that the parameters for a given simulation are such that the differential equation solver can't properly run the model. That usually corresponds to biologically unrealistic parameter settings. We'll ignore them, but if you did a research project and you got such warning or error messages, you'd have to figure out why you get them and only once you fully understand why is it maybe ok to ignore them."
nrec = 1 # number of items to record
out_records = c("Nothing")
out_types = rep("None",nrec)
out_notes = c("")
outcomes = data.frame( TaskID = rep(tid,nrec),
                       TaskText = rep(tasktext,nrec),
                      RecordID = paste0('T',tid,'R',(1:nrec)),
                      Record = out_records, 
                      Type = out_types, 
                      Note = out_notes)
alloutcomes = rbind(alloutcomes,outcomes)
rc = rc + nrec #increment record counter by number of outcomes to record for this task 


#########################
# Task 5
#########################
tid = tid + 1
tasktext = "So far, we only looked at the total variability in outcomes as parameters are being varied/sampled (i.e. the boxplots and summary measures printed underneath the plots). The above approach of exploring the impact of a parameter on results by varying bounds and repeatedly looking at boxplots or summary measures is somewhat tedious. This is where sensitivity analysis comes in. By running the same simulations, but plotting outcomes as scatterplots, we can explore how variation in a paramter leads to variation in outcomes. Let's explore this. Set all input values to their defaults. Then set sample size to 50 and switch the plot type from boxplot to scatterplot. Run the simulation. You'll see how the 3 outcomes of interest vary with one of the varied inputs. Here this is the initial starting value for the bacteria. Which in some sense is not a model parameter, but an initial condition. However, for the purpose of U/S analysis, either is considered a varied input/parameter. Also look at the text below the plots. For each parameter-output pair, the code computes a rank correlation coefficient. Numbers close to 0 mean there is essentially no correlation, close to 1 or -1 means a large positive or negative correlation. (One could compute p-values for these correlations, but they are somewhat meaningless since the values will get smaller the more samples you use, so you can basically produce any p-value you want.). Let's explore this correlation for a parameter we are familiar with from previous apps, the bacteria growth rate. Set all inputs to their defaults, then set samples to 100. Also switch the plot type to _Scatterplot_ and _Parameter for scatterplot_ to $g$. Run the simulation. We found earlier that $B$ at steady state does not depend on $g$, while $I$ does. The correlation coefficients show this as well. Note that they are not exactly 0 and 1 since there is additional variability due to the other parameters being sampled as well."
nrec = 1 # number of items to record
out_records = c("Rank Correlation Coefficient between g and Isteady (2 decimal places).")
out_types = rep("Rounded_Numeric",nrec)
out_notes = rep("Report the value rounded to two significant digits",nrec)
outcomes = data.frame( TaskID = rep(tid,nrec),
                       TaskText = rep(tasktext,nrec),
                      RecordID = paste0('T',tid,'R',(1:nrec)),
                      Record = out_records, 
                      Type = out_types, 
                      Note = out_notes)
alloutcomes = rbind(alloutcomes,outcomes)
rc = rc + nrec #increment record counter by number of outcomes to record for this task 


#########################
# Task 6
#########################
tid = tid + 1
tasktext = "Let's explore in more detail how variability in different parameters impacts results. We'll pretend that we happen to know several model inputs with certainty. To do so, we'll impose no variability for some parameters. For the following parameters, set **both** their lower and upper bound to the specified value: initial value for bacteria and immune response = 1,  _B~max~_ = 1e5,  _d~B~_ = 1, _k_ = r = 10^-4^. Let _d~I~_ vary between 1 and 2 and give $g$ a mean of 5 and variance of 1. 100 samples, seed at 100, scatterplot for $g$. Run the simulation. You'll see a very clear pattern relating the outcomes to $g$. Switch to plotting $d_I$, you'll again see nice patterns. Confirm that - as expected from the steady state equations - that _I~steady~_ depends positively on $g$ and negatively on $d_I$. Both the scatterplots and the correlation coefficient should give you that information. Also note the distribution for _g_ and _d~I~_. The former has more points around its mean and less for lower/higher values, while _d~I~_ values are uniformly distributed along the x-axis. This comes from the underlying assumption about how the parameters are distributed, gamma-distribution versus uniform distribution."
nrec = 1 # number of items to record
out_records = c("Rank Correlation Coefficient between g and Isteady (2 decimal places).")
out_types = rep("Rounded_Numeric",nrec)
out_notes = rep("Report the value rounded to two significant digits",nrec)
outcomes = data.frame( TaskID = rep(tid,nrec),
                       TaskText = rep(tasktext,nrec),
                      RecordID = paste0('T',tid,'R',(1:nrec)),
                      Record = out_records, 
                      Type = out_types, 
                      Note = out_notes)
alloutcomes = rbind(alloutcomes,outcomes)
rc = rc + nrec #increment record counter by number of outcomes to record for this task 

#########################
# Task 7
#########################
tid = tid + 1
tasktext = "In the __Bacteria Model Exploration__ app, we investigated the fact that for some value of $d_I$, we get a switch from a steady state to no steady state (if you don't remember, I suggest you revisit that app first). This behavior can also (accidentally) happen when you sample over parameters. That's not ideal since it means you combine to qualitatively different results in a single distribution. Let's explore this. Reset all values to their defaults. Then set ranges for $d_I$ to be from 1 to 5, 100 samples, a scatterplot with $d_I$ on the x-axis. Also change to plotly as the plot engine. Run the simulation. You should see mostly non-zero values for $I_{steady}$ (you might need to click the auto-scale button on the plot). Now change the range for $d_I$ to be from 10 to 14. You'll now see a large number of zeros for $I_{steady}$, especially for high $d_I$ values. The important take-home message from this is that the influence of a parameter on some outcome can be different over different ranges. For instance in range A-B, the parameter might have a major influence, but once the parameter value goes above B, the parameter does not further influence the result. If you have large uncertainty in your parameters, it might be worth considering both the full range, and dividing the range into smaller areas to see how the parameter behaves."
nrec = 1 # number of items to record
out_records = c("Rank Correlation Coefficient between dI and Isteady for range 10-14 (2 decimal places).")
out_types = rep("Rounded_Numeric",nrec)
out_notes = rep("Report the value rounded to two significant digits",nrec)
outcomes = data.frame( TaskID = rep(tid,nrec),
                       TaskText = rep(tasktext,nrec),
                      RecordID = paste0('T',tid,'R',(1:nrec)),
                      Record = out_records, 
                      Type = out_types, 
                      Note = out_notes)
alloutcomes = rbind(alloutcomes,outcomes)
rc = rc + nrec #increment record counter by number of outcomes to record for this task 
``` 



```{r echo=FALSE}
#save the fully filled task table to a tsv file
alloutcomes$QuizID = paste0(packagename,"_",appsettings$appid)
alloutcomes$AppTitle = appsettings$apptitle
alloutcomes$AppID = appsettings$appid
#remove a few variables from the data frame
savedoutcomes <- dplyr::select(alloutcomes,QuizID,AppID,AppTitle,TaskID,TaskText,RecordID,Record,Type,Note)		
write.table(savedoutcomes, paste0(appsettings$appid,"_tasktable.tsv"), append = FALSE, sep = "\t", row.names = F, col.names = TRUE)
```


```{r, echo=FALSE, results='asis'}
# Take all the text stored in the table and print the tasks and items to record
write_tasktext(alloutcomes)
```







## Further Information {#shinytab4}
* For this app, the underlying function running the simulation is called ``r appsettings$simfunction``. That function repeatedly calls ``r appsettings$underlying_function``. 
* This app (and all others) are structured such that the Shiny part (the graphical interface you are using) calls one or several underlying R functions which run the simulation for the model of interest and return the results. You can call them directly, without going through the shiny app. Use the `help()` command for more information on how to use the functions directly. If you go that route, you need to use the results returned from this function and produce useful output (such as a plot) yourself. 
* You can also download all simulator functions and modify them for your own purposes.  Of course to modify these functions, you'll need to do some coding.
* For examples on using the simulators directly and how to modify them, read the package vignette by typing `vignette('DSAIRM')` into the R console.
* Good papers explaining uncertainty and sensitivity analysis in a bit more detail are [@hoare08; @marino08]. 


### References


